---
title: "Tuning context length with rolling-origin resampling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tuning context length with rolling-origin resampling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

How much history should a foundation model condition on? More context is not
always better — it can dilute recent signal and cost compute. Because `tsfm`
exposes `context_length` as a tunable `dials` parameter, you can answer this
empirically with the standard tidymodels tuning machinery over **rolling-origin**
(time-aware) resamples, rather than random cross-validation that would leak the
future into the past.

The code below uses the executable `stub` fixture to demonstrate the mechanics.
It is not a foundation-model benchmark; repeat the workflow with TimesFM only
after that checkpoint passes the package's support gates.

## The spec

Mark `context_length` for tuning; everything else is fixed through the engine:

```{r}
library(tsfm)
library(parsnip)
library(workflows)
library(tune)
library(dials)
library(rsample)

spec <- tsfm_reg(context_length = tune()) |>
  set_engine("tsfm",
             model_id = "stub",
             index    = "date",
             id       = "store")

wf <- workflow() |>
  add_formula(demand ~ date) |>
  add_model(spec)
```

## Rolling-origin resamples

`rsample::sliding_period()` builds analysis/assessment splits that always move
forward in time — the assessment window is strictly after its analysis window:

```{r}
folds <- sliding_period(
  train_df,
  index  = date,
  period = "month",
  lookback   = Inf,   # expanding-origin: use all history up to each split
  assess_stop = 1     # score one month ahead
)
```

## Search

Give `context_length` a range and let `tune_grid()` score each candidate with a
proper (quantile-aware) metric:

```{r}
grid <- grid_regular(
  context_length(range = c(64L, 512L)),
  levels = 5
)

res <- tune_grid(
  wf,
  resamples = folds,
  grid      = grid,
  metrics   = metric_set(yardstick::rmse, yardstick::mae)
)

show_best(res, metric = "rmse")
best <- select_best(res, metric = "rmse")

final_fit <- finalize_workflow(wf, best) |>
  fit(data = train_df)
```

## Notes

- `context_length` bounds the most recent history retained by each fitted
  engine; the end-to-end behavior is covered with the stub in package tests.
- The heavy per-candidate cost is inference, not training (these are zero-shot
  models), so keep the grid small and prefer `sliding_period()` windows that
  reflect your real forecast horizon.
