---
title: "Zero-shot forecasting in a workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Zero-shot forecasting in a workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`tsfm` gives every time-series foundation model one uniform, tidymodels-native
interface. This vignette shows the two ways to forecast a panel of series, and
the property that makes the package worth having: **swapping one foundation
model for another means changing a single argument.**

## Load a model

`tsfm_pretrained()` resolves a model from the Hugging Face Hub (revision pinned)
and returns a uniform handle. The same call works for every architecture:

```{r}
library(tsfm)

model <- tsfm_pretrained("amazon/chronos-2")
```

Before doing any work, ask what the model can do. The capability report is
checked *before* inference, so an unsupported request fails fast with a clear
message instead of deep inside a tensor op:

```{r}
tsfm_capabilities(model)
#> <tsfm_capabilities>
#>   architecture:      chronos2
#>   max_context:       8192        quantiles: native
#>   multivariate:      FALSE       samples:   no
#>   past_covariates:   TRUE        future_covariates: TRUE
#>   static_covariates: FALSE       fine_tunable: FALSE
#>   license:           Apache-2.0
```

## Path 1 — the `forecast()` convenience surface (tsibble)

Hand a panel of history and a horizon; get a forecast object back that converts
to a `fable` for the tidyverts accuracy/reconciliation stack:

```{r}
library(tsibble)

fc <- forecast(model, tourism_history, h = 12,
               quantile_levels = c(0.1, 0.5, 0.9))

fc |>
  as_fable() |>
  fabletools::accuracy(tourism_actuals)
```

## Path 2 — the tidymodels workflow (canonical)

The canonical path is a `parsnip` spec inside a `workflow()`. The model is chosen
by the engine's `model_id`; **everything else stays the same.**

```{r}
library(parsnip)
library(workflows)

spec <- tsfm_reg() |>
  set_engine("tsfm",
             model_id = "amazon/chronos-2",
             index    = "date",
             id       = "store")

wf <- workflow() |>
  add_formula(demand ~ date) |>
  add_model(spec)

fit <- fit(wf, data = train_df)
predict(fit, new_data = future_df)   # tibble: .pred, .pred_lower, .pred_upper, ...
```

### Swapping models is a one-line change

To forecast with TTM instead of Chronos-2, edit only `model_id` — the R analogue
of Darts' "change only the model name":

```{r}
spec_ttm <- tsfm_reg() |>
  set_engine("tsfm",
             model_id = "ibm-granite/granite-timeseries-ttm-r2",  # <- only change
             index    = "date",
             id       = "store")
```

## Tuning the context length

`context_length` is a tunable `dials` parameter, so a rolling-origin search over
`rsample` resamples works out of the box:

```{r}
library(tune)
library(rsample)

spec <- tsfm_reg(context_length = tune()) |>
  set_engine("tsfm", model_id = "amazon/chronos-2", index = "date", id = "store")

tune_grid(
  workflow() |> add_formula(demand ~ date) |> add_model(spec),
  resamples = sliding_period(train_df, date, "month"),
  grid      = 5
)
```

## Status

`amazon/chronos-2` is served today through a `brulee`-backed adapter. The native
TTM and TimesFM implementations follow on the roadmap; because the loader,
capability, batching, forecast, and parsnip layers shown here are model-agnostic,
those arrive as new architectures behind the *same* code above.
